---
title: "Database Quality Report"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
params:
  db_data: !r data.frame()
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)
library(ggplot2)
library(dplyr)

# Load the data passed from the Shiny app
df <- params$db_data
```

# Executive Summary

This report summarises the current state of the patient database. It was generated on **`r Sys.Date()`** and contains **`r nrow(df)`** patient records with **`r ncol(df)`** variables.

---

# Database Overview

```{r overview_table}
# Basic info about the dataset
overview <- data.frame(
  Metric = c("Total Records", "Total Variables", "Report Generated"),
  Value  = c(nrow(df), ncol(df), as.character(Sys.Date()))
)
kable(overview, caption = "General Database Statistics")
```

## First rows of data

```{r head_table}
kable(head(df, 10), caption = "First 10 records in the database")
```

---

# Quality Metric 1: Completeness (Missing Values)

We check how many values are missing (`NA`) in each variable. Variables with high missing percentages may need attention.

```{r completeness_table}
missing_pct <- sapply(df, function(x) round(sum(is.na(x)) / nrow(df) * 100, 2))
completeness_df <- data.frame(
  Variable           = names(missing_pct),
  Missing_Count      = sapply(df, function(x) sum(is.na(x))),
  Missing_Percentage = missing_pct,
  row.names          = NULL
)
kable(completeness_df, caption = "Missing values per variable")
```

```{r completeness_plot, fig.width=8, fig.height=5}
# Red bars = variables with missing data, blue = complete
ggplot(completeness_df, aes(x = reorder(Variable, Missing_Percentage),
                             y = Missing_Percentage,
                             fill = Missing_Percentage > 0)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "tomato"),
                    labels = c("Complete", "Has Missing"),
                    name   = "Status") +
  theme_minimal(base_size = 13) +
  labs(title = "Missing Data (%) per Variable",
       x = "Variable", y = "Missing (%)")
```

**Interpretation:** Variables shown in red have at least one missing value and should be reviewed.

---

# Quality Metric 2: Consistency (Sex Variable)

We check whether the `Sex` variable only contains expected values (Male, Female, Other). Any other values indicate data entry errors.

```{r consistency}
if ("Sex" %in% colnames(df)) {
  sex_table <- as.data.frame(table(df$Sex, useNA = "ifany"))
  colnames(sex_table) <- c("Value", "Count")
  sex_table$Percentage <- round(sex_table$Count / sum(sex_table$Count) * 100, 2)

  standard_values <- c("Male", "Female", "Other", "m", "f", "M", "F")
  non_standard <- sex_table[!sex_table$Value %in% standard_values & !is.na(sex_table$Value), ]

  kable(sex_table, caption = "Distribution of the Sex variable (including NAs)")
} else {
  cat("Variable 'Sex' not found in the database.")
}
```

```{r consistency_alert, results='asis'}
# Show a pass/fail message depending on whether non-standard values exist
if ("Sex" %in% colnames(df)) {
  if (nrow(non_standard) > 0) {
    cat("> **WARNING:** Non-standard values found in Sex:",
        paste(non_standard$Value, collapse = ", "), "\n")
  } else {
    cat("> **PASS:** All values in Sex are within the expected range.\n")
  }
}
```

---

# Quality Metric 3: Accuracy (Dosage Outliers)

We use the IQR method to detect extreme dosage values that may indicate data entry mistakes or unusual cases. Values below Q1 − 1.5×IQR or above Q3 + 1.5×IQR are flagged as outliers.

```{r accuracy_plot, fig.width=7, fig.height=5}
if ("Dosage_mg" %in% colnames(df)) {
  df$Dosage_mg <- suppressWarnings(as.numeric(df$Dosage_mg))

  ggplot(df, aes(y = Dosage_mg)) +
    geom_boxplot(fill = "tomato", color = "darkred",
                 outlier.color = "black", outlier.size = 3) +
    theme_minimal(base_size = 13) +
    labs(title = "Dosage Distribution with Outliers",
         y = "Dosage (mg)") +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
} else {
  cat("Variable 'Dosage_mg' not found in the database.")
}
```

```{r accuracy_table}
if ("Dosage_mg" %in% colnames(df)) {
  df$Dosage_mg <- suppressWarnings(as.numeric(df$Dosage_mg))
  q1  <- quantile(df$Dosage_mg, 0.25, na.rm = TRUE)
  q3  <- quantile(df$Dosage_mg, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower_bound <- q1 - 1.5 * iqr
  upper_bound <- q3 + 1.5 * iqr

  outliers <- df[!is.na(df$Dosage_mg) &
                   (df$Dosage_mg < lower_bound | df$Dosage_mg > upper_bound), ]

  # Summary table of dosage statistics
  stats_df <- data.frame(
    Statistic = c("Minimum", "Q1 (25th pct)", "Median", "Q3 (75th pct)",
                  "Maximum", "IQR", "Lower fence", "Upper fence", "Outlier count"),
    Value = c(min(df$Dosage_mg, na.rm = TRUE), q1,
              median(df$Dosage_mg, na.rm = TRUE), q3,
              max(df$Dosage_mg, na.rm = TRUE), iqr,
              lower_bound, upper_bound, nrow(outliers))
  )
  kable(stats_df, caption = "Dosage summary statistics")

  if (nrow(outliers) > 0) {
    kable(head(outliers, 20), caption = "Records flagged as dosage outliers (first 20)")
  }
}
```

---

# Audit Log

> **Note:** A complete audit trail (user, timestamp, action) is stored in the `audit_logs` table of the SQLite database. It is not included in this report to protect user privacy.

---

# Conclusions

```{r conclusions, results='asis'}
total    <- nrow(df)
any_miss <- any(sapply(df, function(x) sum(is.na(x))) > 0)

# Count dosage outliers
dose_flag <- if ("Dosage_mg" %in% colnames(df)) {
  d  <- suppressWarnings(as.numeric(df$Dosage_mg))
  q1 <- quantile(d, 0.25, na.rm = TRUE)
  q3 <- quantile(d, 0.75, na.rm = TRUE)
  sum(d < q1 - 1.5 * (q3 - q1) | d > q3 + 1.5 * (q3 - q1), na.rm = TRUE)
} else 0

cat(sprintf("- **Total records analysed:** %d\n", total))
cat(sprintf("- **Missing values present:** %s\n",
    if (any_miss) "Yes – review required" else "No – database is complete"))
cat(sprintf("- **Dosage outliers detected:** %d record(s)\n", dose_flag))
cat("\nThese findings should be reviewed by the data manager before any further analysis.\n")
```

---

*Report generated automatically by the Biomedical Data Hub · `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`*
